\systemEnding: Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 paper.tex
miscellaneous edits

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence. Recent work has successfully used
symbolic search techniques (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}), neural networks trained from
a corpus of examples (e.g., RobustFill:~\cite{devlin2017robustfill},
DeepCoder:~\cite{balog2016deepcoder}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search~\cite{ngds}) to
synthesize programs for task domains such as string transformations,
list processing, and robot navigation and planning. However, all these
approaches -- symbolic, neural and neural-symbolic -- rely upon a
hand-engineered \emph{Domain Specific Language} (DSL). The DSL is an
inventory of restricted programming primitives, encoding domain
specific knowledge about the space of programs. In
practice
we often have only a few input/output examples for each
program to be induced, and thus success often hinges on having a good
DSL, which
provides a crucial inductive bias for what would otherwise be an
unconstrained search through the space of all computable functions.
Here we ask, to what extent can we dispense with such highly
hand-engineered DSLs?

@@

Our algorithm is called \system because it is based on a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve each of the tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL. The DSL
distills commonalities across programs that solve the tasks, letting
the agent solve other related program learning problems. The neural
network ensures that searching for program solutions remains tractable
even as the DSL (and hence the search space for programs) expands.

@@

Our goal  is to induce a DSL while finding programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) searching for programs; (2) learning a neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation is that these steps bootstrap off each other (Fig.~\ref{feeding}):
\\\noindent \textbf{Wake: Searching for programs.}  Our program search  is informed by both the DSL and the recognition model.
When these improve we find better programs.
\\\noindent\textbf{Sleep-R: Learning a recognition model.} The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.
\\\noindent\textbf{Sleep-G: Improving the DSL.} We induce a DSL from the programs we have found so far which solve the tasks;
  as we solve more tasks, we can hone in on richer DSLs that more closely match the domain.

@@

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:
(1) A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the search algorithm is very simple.
(2) Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
  the space of of programs.

@@

Evaluating $\mathcal{L}_{\text{HM}}$ involves sampling programs from
the current DSL, running them to get their outputs,
and then training $q$ to regress from the outputs to the program.
Since these programs map inputs to outputs,
we need some way of sampling these inputs as well.
Our solution to this problem is to sample the inputs
from the empirical observed distribution of inputs in $X$.

@@

\section{Experiments}
We first present our experimental results qualitatively,
illustrating the DSLs we
learn.
We then show quantitative results (Sec.~\ref{quantitative})
comparing \system with lesioned versions
of our model that are analogous to prior work.

@@

We initially give our system only addition and multiplication,
along with the possibility of introducing real-valued parameters, which we write as $\mathcal{R}$.
We define the likelihood of an observation $x$ by assuming a Gaussian noise model for the input/output examples,
and penalize continuous degrees of freedom using the BIC~\cite{Bishop:2006:PRM:1162264}.
%% \begin{align*}
%%   \log \probability[x|p] \triangleq& \sum_{n\leq N}\log
%%   \mathcal{N}(p(i_n,\vec{\mathcal{R}}^*)|o_n) - \frac{D\log N}{2}
%% \end{align*}

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom:
for example, it represents the linear function $-x+2$ with the program 
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)}, which has two continuous degrees of freedom -- while representing quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}.
This falls out of our Bayesian framing -- both the bias towards shorter programs (more likely under $\mathcal{D},\theta$) and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
  $f_0\,=\,$\code{(+ $\mathcal{R}$)}\\
  $f_1($\code{x}$)\,=\,$\code{(* (* (+ x $\mathcal{R}$) $\mathcal{R}$) x)} \\
  $f_2($\code{x}$)\,=\,$\code{(+ $\mathcal{R}$ ($f_1$ x))} \\
  $f_3($\code{x}$)\,=\,$\code{(* x ($f_2$ x))}\\
  $f_4($\code{x}$)\,=\,$\code{($f_0$ ($f_3$ x))}\\
  $f_5($\code{x}$)\,=\,$\code{(* x ($f_4$ x))}\\
  $f_6($\code{x}$)\,=\,$\code{($f_0$ ($f_4$ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some induced DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns to build 4th order polynomials.}\label{regressionDSL}
\end{table}

@@

\begin{table}[t]
  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \code{0}, \code{+1},\code{-1},     \code{""},    \code{","},     \code{" "},    \code{"<"}, \code{">"}  \\
    \code{split}, \code{join}, \code{map}, 
    \code{+}, \code{chr->str}, \code{trim}, \code{nth}\\
    \code{length}, \code{slice},    \code{lower}, \code{upper}, \code{capitalize}
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\code{c}$) \,=\, $\code{(+ (chr->str c))}\\
      $f_1($\code{i$,$c$,$s}$) \,=\, $\code{(+ (nth i (split c s))}\\
        \phantom{$f_1($\code{i$,$c$,$s}$) \,=\, $\code{(+ }}\code{(chr->str c))}\\
      $f_2($\code{s$,$c}$) \,=\, $\code{((}$f_0$\code{ c) (}$f_1$\code{ (1+ 0) c s))} \\
        \hspace{0.5cm}($f_2$: \emph{Extract substring delimited by character})\\
      $f_3($\code{c$,$s}$) \,=\, $\code{(nth (1+ 0) (split c (nth 0 s)))}\\
        \hspace{0.5cm}($f_3$: \emph{2nd substring delimited by character})
    %% $f_0($\code{i,c,s}$) = $    \code{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\code{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}

@@

\begin{table}
\centering
\begin{tabular}{lll}
  \toprule
  Name & Input & Output \\\midrule
  append-4 & [7\, 0\, 2] & [7\, 0\, 2\, 4] \\
  len & [3\, 5\, 12\, 1] & 4 \\
  has-2 & [4\, 5\, 7\, 4] & \code{false} \\
  repeat-2 & [7\, 0] & [7\, 0\, 7\, 0] \\
  drop-3 & [0\, 3\, 8\, 6\, 4] & [6\, 4] \\
  count-head-in-tail & [1\, 2\, 1\, 1\, 3] & 2 \\
  rotate-2 & [8\, 14\, 1\, 9] & [1\, 9\, 8\, 14] \\
  pow-3 & [10\, 4\, 13] & [1000\, 64\, 2197] \\
  keep-mod-5 & [5\, 9\, 14\, 6\, 3\, 0] & [5\, 0] \\
  \bottomrule
\end{tabular}
\caption{Sample tasks from our list function domain. %% We used 15 input/output
  %% examples for each task to reduce ambiguity in the task's function,
  %% ensuring solutions achieve the desired concept.
}
\label{listexamples}
\end{table}

\begin{table}[t]
  \centering
  \begin{tabular}{l}
    \toprule
    $f_0($\code{i$,\ell$}$) \,=\, $\code{(singleton (index i $\ell$))}\\
    \hspace{0.5cm}($f_0$: \emph{Put the $i$-th index into a list})\\
    $f_1($\code{i$,\ell$}$) \,=\, $\code{(++ $\ell$ ($f_0$ i $\ell$))}\\
    \hspace{0.5cm}($f_1$: \emph{Append the $i$-th index})\\
    $f_2($\code{n$,\ell$}$) \,=\, $\code{(any (lambda (x) (= n x)) $\ell$)}\\
    \hspace{0.5cm}($f_1$: \emph{Whether $n$ appears in the list})\\
    $f_3($\code{i$,\ell$}$) \,=\, $\code{(index (negate i) (sort $\ell$))}\\
    \hspace{0.5cm}($f_3$: \emph{Get the $i$-th largest number})\\
    $f_4($\code{n$,\ell$}$) \,=\, $\code{(mapi (lambda (i x) (mod x n)) $\ell$)}\\
    $f_5($\code{k$,\ell$}$) \,=\, $\code{(mapi (lambda (i x) (+ x k)) $\ell$)}\\
    $f_6($\code{k$,$n$,\ell$}$) \,=\, $\code{($f_4$ n ($f_5$ k $\ell$)}\\
    \hspace{0.5cm}($f_6$: \emph{Caesar shift of $k$ in integers modulo $n$})\\
  \bottomrule
  \end{tabular}
  \caption{Some induced primitives for list functions.}\label{listinduced}
\end{table}

>>>>

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence. Recent work has successfully used
symbolic search techniques (e.g., Metagol:~\cite{muggleton2015meta},
FlashFill:~\cite{gulwani2011automating}), neural networks trained from
a corpus of examples (e.g., RobustFill:~\cite{devlin2017robustfill},
DeepCoder:~\cite{balog2016deepcoder}), and hybrids of neural and
symbolic methods (e.g., Neural-guided deductive search~\cite{ngds}) to
synthesize programs for task domains such as string transformations,
list processing, and robot navigation and planning. However, all these
approaches -- symbolic, neural and neural-symbolic -- rely upon a
hand-engineered \emph{Domain Specific Language} (DSL). The DSL is an
inventory of restricted programming primitives, encoding domain
specific knowledge about the space of programs. In
practice
we often have only a few input/output examples for each
program to be induced, and thus success often hinges on having a good
DSL, which
provides a crucial inductive bias for what would otherwise be an
unconstrained search through the space of all computable functions.
Here we ask, to what extent can we dispense with such highly
hand-engineered domain-specific languages?

@@

Our algorithm is called \system because it is based on a
novel kind of ``wake-sleep'' learning (c.f. \cite{hinton1995wake}), iterating
between ``wake'' and ``sleep'' phases to achieve three goals: finding
programs that solve each of the tasks; creating a DSL by discovering
and reusing domain-specific subroutines; and training a neural network
that efficiently guides search for programs in the DSL. The DSL
distills commonalities across programs that solve the tasks, letting
the agent solve other related program induction problems. The neural
network ensures that searching for programs remains tractable
even as the DSL (and hence the search space for programs) expands.

@@

Our goal  is to induce a DSL while finding programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) searching for programs; (2) learning a neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation is that these steps bootstrap off each other (Fig.~\ref{feeding}):
\\\noindent \textbf{Wake: Searching for programs.}  Our program search  is informed by both the DSL and the recognition model.
When these improve we find more programs solving the tasks.
\\\noindent\textbf{Sleep-R: Learning a recognition model.} The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.
\\\noindent\textbf{Sleep-G: Improving the DSL.} We induce a DSL from the programs we have found so far which solve the tasks;
  as we solve more tasks, we can hone in on richer DSLs that more closely match the domain.

@@

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:
(1) A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the search algorithm is very simple.
(2) Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
  the space of  programs.

@@

Evaluating $\mathcal{L}_{\text{HM}}$ involves sampling programs from
the current DSL, running them to get their outputs,
and then training $q$ to regress from the input/outputs to the program.
Since these programs map inputs to outputs,
we need to sample the inputs as well.
Our solution is to sample the inputs
from the empirical observed distribution of inputs in $X$.

@@

\section{Experiments}
We first present our experimental results qualitatively,
illustrating the learned DSLs.
We then show quantitative results (Sec.~\ref{quantitative})
comparing \system with lesioned versions
of our model that are analogous to prior work.

@@

We initially give our system only addition and multiplication,
along with the possibility of introducing real-valued parameters, which we write as $\mathcal{R}$.
We define the likelihood of an observation $x$ by assuming a Gaussian noise model for the input/output examples,
and penalize the use of continuous parameters using the BIC~\cite{Bishop:2006:PRM:1162264}.
%% \begin{align*}
%%   \log \probability[x|p] \triangleq& \sum_{n\leq N}\log
%%   \mathcal{N}(p(i_n,\vec{\mathcal{R}}^*)|o_n) - \frac{D\log N}{2}
%% \end{align*}

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom:
for example, it represents the linear function $-x+2$ with the program 
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)}, which has two continuous degrees of freedom -- while representing quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}.
This falls out of our Bayesian framing -- both the bias towards shorter programs (more likely under $\mathcal{D},\theta$) and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
  $f_0\,=\,$\code{(+ $\mathcal{R}$)}\\
  $f_1($\code{x}$)\,=\,$\code{(* (* (+ x $\mathcal{R}$) $\mathcal{R}$) x)} \\
  $f_2($\code{x}$)\,=\,$\code{(+ $\mathcal{R}$ ($f_1$ x))} \\
  $f_3($\code{x}$)\,=\,$\code{(* x ($f_2$ x))}\\
  $f_4($\code{x}$)\,=\,$\code{($f_0$ ($f_3$ x))}\\
  $f_5($\code{x}$)\,=\,$\code{(* x ($f_4$ x))}\\
  $f_6($\code{x}$)\,=\,$\code{($f_0$ ($f_4$ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some learned DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns to build 4th order polynomials.}\label{regressionDSL}
\end{table}

@@

\begin{table}[t]
  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \code{0}, \code{+1},\code{-1},     \code{""},    \code{","},     \code{" "},    \code{"<"}, \code{">"}  \\
    \code{split}, \code{join}, \code{map}, 
    \code{+}, \code{chr->str}, \code{trim}, \code{nth}\\
    \code{length}, \code{slice},    \code{lower}, \code{upper}, \code{capitalize}
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\code{c}$) \,=\, $\code{(+ (chr->str c))}\\
      $f_1($\code{i$,$c$,$s}$) \,=\, $\code{(+ (nth i (split c s))}\\
        \phantom{$f_1($\code{i$,$c$,$s}$) \,=\, $\code{(+ }}\code{(chr->str c))}\\
      $f_2($\code{s$,$c}$) \,=\, $\code{((}$f_0$\code{ c) (}$f_1$\code{ (1+ 0) c s))} \\
        \hspace{0.5cm}($f_2$: \emph{Extract substring delimited by character})\\
      $f_3($\code{c$,$s}$) \,=\, $\code{(nth (1+ 0) (split c (nth 0 s)))}\\
        \hspace{0.5cm}($f_3$: \emph{2nd substring delimited by character})
    %% $f_0($\code{i,c,s}$) = $    \code{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\code{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some learned DSL primitives.}\label{textPrimitives}
  \end{table}

@@

\begin{table}
\centering
\begin{tabular}{lll}
  \toprule
  Name & Input & Output \\\midrule
  append-4 & [7\, 0\, 2] & [7\, 0\, 2\, 4] \\
  len & [3\, 5\, 12\, 1] & 4 \\
  has-2 & [4\, 5\, 7\, 4] & \code{false} \\
  repeat-2 & [7\, 0] & [7\, 0\, 7\, 0] \\
  drop-3 & [0\, 3\, 8\, 6\, 4] & [6\, 4] \\
  count-head-in-tail & [1\, 2\, 1\, 1\, 3] & 2 \\
  rotate-2 & [8\, 14\, 1\, 9] & [1\, 9\, 8\, 14] \\
  pow-3 & [10\, 4\, 13] & [1000\, 64\, 2197] \\
  keep-mod-5 & [5\, 9\, 14\, 6\, 3\, 0] & [5\, 0] \\
  \bottomrule
\end{tabular}
\caption{Sample tasks from our list function domain %% We used 15 input/output
  %% examples for each task to reduce ambiguity in the task's function,
  %% ensuring solutions achieve the desired concept.
}
\label{listexamples}
\end{table}

\begin{table}[t]
  \centering
  \begin{tabular}{l}
    \toprule
    $f_0($\code{i$,\ell$}$) \,=\, $\code{(singleton (index i $\ell$))}\\
    \hspace{0.5cm}($f_0$: \emph{Put the $i$-th index into a list})\\
    $f_1($\code{i$,\ell$}$) \,=\, $\code{(++ $\ell$ ($f_0$ i $\ell$))}\\
    \hspace{0.5cm}($f_1$: \emph{Append the $i$-th index})\\
    $f_2($\code{n$,\ell$}$) \,=\, $\code{(any (lambda (x) (= n x)) $\ell$)}\\
    \hspace{0.5cm}($f_1$: \emph{Whether $n$ appears in the list})\\
    $f_3($\code{i$,\ell$}$) \,=\, $\code{(index (negate i) (sort $\ell$))}\\
    \hspace{0.5cm}($f_3$: \emph{Get the $i$-th largest number})\\
    $f_4($\code{n$,\ell$}$) \,=\, $\code{(mapi (lambda (i x) (mod x n)) $\ell$)}\\
    $f_5($\code{k$,\ell$}$) \,=\, $\code{(mapi (lambda (i x) (+ x k)) $\ell$)}\\
    $f_6($\code{k$,$n$,\ell$}$) \,=\, $\code{($f_4$ n ($f_5$ k $\ell$)}\\
    \hspace{0.5cm}($f_6$: \emph{Caesar shift of $k$ in integers modulo $n$})\\
  \bottomrule
  \end{tabular}
  \caption{Some learned DSL primitives for list functions}\label{listinduced}
\end{table}
